Estimating duration in the presence of misclassification

### 2024-02-08 transmission.tex

In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
This estimation is based only on the CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
I will show that this data is sufficient to estimate the parameters of the transmission process, suggesting surveys of this type are sufficient for surveillance of an epidemic.

I take two approaches; the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
In addition to estimating incidence, this approach also estimates the components of transmission, \eg whether changes can be explained by changes in contact rates.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

\todo[inline]{
    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Toffe?
    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
}
\section{Data} \label{transmission:sec:data}

\todo[inline]{This is now in \cref{E-biology-data:sec:posivity})}

Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
\begin{enumerate}
    \item $[0, 11)$: children of pre-school and primary school age.
    \item $[11, 16)$: children of compulsory secondary-school age.
    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
    \item $[25, 50)$: younger working age adults.
    \item $[50, 70)$: older working age adults.
    \item $[70, \infty)$: elderly.
\end{enumerate}
These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.

I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.

I will model the test results as overdispersed relative to a binomial model.
Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}. 
The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
$\rho=0$ means the beta-binomial coincides with the binomial distribution.
As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

A deterministic relationship between true prevalence an incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
A well-established methodology, using multilevel regression and poststratification (MRP)\todo{cite MRP}, has already been developed for estimating prevalence from the CIS data\todo{cite Koen's stuff}.
Therefore, I extend the methodology previously used to estimate prevalence to estimate incidence.

In MRP, a Bayesian multilevel logistic regression model is first fit to predict RT-PCR positivity from demographic covariates.
Then, the posterior distribution is used to predict the prevalence in each stratum.
Finally, the prevalence is poststratified (as explained later) to the CIS target population and subpopulations of interest.
I modify this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

The deterministic relationship between incidence proportion and prevalence means that either can be inferred, with the other then calculated.
Denote the quantity to be inferred at time $t$ is strata $i$ as $\pi_{it}$.
Note that for any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
The statistical model and inference procedure, including the borrowing of information, should have the following properties.
\begin{itemize}
    \item The smoothing model should penalise towards exponential growth.
        Penalising towards exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a single time unit (day) which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:positivity}).
    \item Computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see\todo{ref section on SRS computational limitations}).
        In particular, MCMC should be avoided because incorporating sufficient covariates to account for all groups led to a model that was too computationally expensive~\citePersonalComms{Koen Pouwels}.
\end{itemize}

\todo[inline]{I am struggling the framing of these paragraphs. Maybe talk about being linear in its parameters and pre-existing software packages and not mention GLMs?}
The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM} or Gaussian Markov Random Field (GMRF).
Standard, efficient, and widely-used software can then be applied for inference.

To use a GLM or GMRF, express incidence proportion or prevalence in a stratum $i$ on day $t$, $\pi_{it}$, as $\logit(\pi_{it}) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
If $g$ is chosen appropriately, this is a GLM with a logit link function.

The logit transformation serves two purposes.
First, it ensures that the incidence proportion or prevalence is between 0 and 1.
Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_{it})$ over time implies approximately exponential growth of $\pi_{it}$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
Penalising towards exponential growth is a natural formulation because mechanistic models predict expoential growth over short periods of time.
Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.

However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not expressible as a GLM.
Specifically, the prevalence (or an element-wise transformation of it) is no longer linear in the model parameters.
The relationship between incidence and prevalence is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}.
Therefore, the relationship between the logit of incidence proportion and prevalence is non-linear.

Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC~\citePersonalComms{Koen Pouwels}.
Instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used\todo{cite INLA}.

I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
First, a Gaussian Markov Random Field (GMRF) logistic regression model is used to estimate the prevalence in each stratum.
For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
\begin{align}
    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho)
\end{align}
The terms are as follows.
\begin{itemize}
    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
    \item $\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity. $\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively. Their standard deviation have a penalised complexity prior\todo{cite PC prior} with a 10\% probability of being greater than 1.
    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{at} \dist \N(0, \sigma_l^2)$.
    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$. The likelihood is a beta-binomial model (see discussion in \cref{transmission:sec:data}). an overdispersed binomial. 
\end{itemize}
All parameters have the default R-INLA priors, except where stated above.
For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.

@@

Having fit the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is $P_t = \E(P_t \mid \vec{Z})$.
\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.

The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
This gives a lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
$\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
% Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
% Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.

@@

Model misspecification means that the procedure described could produce negative incidence.
\todo{This paragraph might be too terse.}
For any stratum and day $t$, this occurs if the prevalence reduces between day $t$ and $t+1$ more than the number of recoveries on day $t$, based on the incidence up until day $t$ and the survival function.

There are two places where this misspecification could occur.
The first is the survival matrix, $\matr{S}$.
For example, \cref{E-imperf-test} showed that it is sensitive to assumptions made, such as the assumed $\psens$ and ....
The second is the specification of the statistical model used to estimate prevalence, for example the form of the smoothing used.

\todo[inline]{Maybe relate following paragraph to the GLM discussion at the start of the section - its basically the same point}
A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
% MCMC is likely too computationally expensive.
% Other approaches, such as a Laplace approximation or variational inference, could be considered.
For this analysis, I replaced any value of incidence estimated to be negative with 0.

@@

During November, incidence starts rising, although starting at different times in different regions.
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
The distribution of infections among age groups remaining constant over short periods of time is the pattern predicted by standard age-stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) in equilibrium, if behavioural changes are independent of age group, and there is negligible impact of immunity.
The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

@@

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probability upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

>>>>

In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England, solely on the basis of .
CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
I will show that the CIS data is sufficient to derive such estimates, suggesting surveys of this type would be sufficient for surveillance of an epidemic.

I take two approaches, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}. This approach enables estimation of incidence within (sub)populations of England. Then, in \cref{SEIR}, I use a mechanistic compartmental model, a variation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}, to estimate both incidence and components of transmission, \eg and whether changes in transmission can be explained by changes in contact rates.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

\todo[inline]{
    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Taffe?
    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
}
\section{Data} \label{transmission:sec:data}

\todo[inline]{This is now in the final section of chapter 2.}

% Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
% \begin{enumerate}
%     \item $[0, 11)$: children of pre-school and primary school age.
%     \item $[11, 16)$: children of compulsory secondary-school age.
%     \item $[16, 25)$: adolescents and young adults who may remain in formal education.
%     \item $[25, 50)$: younger working age adults.
%     \item $[50, 70)$: older working age adults.
%     \item $[70, \infty)$: elderly.
% \end{enumerate}
% These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.

% I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
% This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.

% I will model the test results as overdispersed relative to a binomial model.
% Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}. 
% The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
% $\rho=0$ means the beta-binomial coincides with the binomial distribution.
% As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
The deterministic relationship between incidence proportion and prevalence means that either can be inferred, with the other then calculated.
Estimation of prevalence from the CIS data has been carried out through a well-established methodology, used throughout the pandemic~\autocite{pouwelsCommunity,cisMethodsONS}.
The method is based on multilevel regression and post-stratification (MRP)\todo{cite MRP}.
In this section, I will extend this methodology to also estimate incidence.

First, I will motivate this approach, as opposed to direct estimation of incidence.
Denote the quantity to be inferred (incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
For any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
The statistical model and inference procedure should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise towards exponential growth. Penalising towards exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time. Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:positivity}).
    \item Borrow information across the different covariates.
        The borrowing should penalise towards the distribution of incidence/prevalence across strata in the same geographic region being constant.
        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see\todo{ref section on SRS computational limitations}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function penalised towards being linear in $t$.
The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$.
Combined with property 4, the form of $g$ should be penalised towards $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified.
However, it should be possible for sufficient information in the data to move the posterior estimate of $\pi_{it}$ away from this model.

Properties 3 and 4 suggest the extension of the previously-developed MRP method.
% This method has previously been used successfully to estimate prevalence, corrected for non-response bias, from the CIS data~\autocite{cisMethodsONS,pouwelsCommunity,pouwelsMRPvaccination}.
In MRP, a Bayesian multilevel logistic regression model is used to predict RT-PCR test results based on demographic covariates.
Then, the posterior distribution is used to predict the prevalence in each stratum.
Finally, the prevalence is post-stratified to the CIS target population and sub-populations of interest.
I propose this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
Further details on this methodology follow.

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, MCMC should be avoided.
Incorporating sufficient covariates to adjust for the non-response bias over a period of more than 6--8 weeks leads to a model that was too computationally expensive~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)\todo{cite INLA}, implemented in R-INLA\todo{cite INLA}.

The use of R-INLA requires that $\pi_{i,t}$ represent incidence proportion rather than prevalence.
If $\vec{\pi}$ represented incidence proportion, then $\logit \vec{pi}$ could be linear in the model parameters, as explained previously, while satisfying the other criteria.
However, the expected prevalence at time $t$ is linear in $\vec{\pi}$ (see \cref{E-inc-prev:eq:EPt}), and hence non-linear in $\logit \vec{pi}$.
The expected proportion of positive tests (\ie the expectation of the data likelihood) is equal to the prevalence.
Applying a non-linear transformation then a linear transformation in this way to form the data likelihood's expectation cannot be implemented within R-INLA's framework.

I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
First, a Gaussian Markov Random Field (GMRF) logistic regression model is used to estimate the prevalence in each stratum.
For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
\begin{align}
    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho)
\end{align}
The terms are as follows.
\begin{itemize}
    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
    \item $\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity. $\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively. Their standard deviation have a penalised complexity prior\todo{cite PC prior} with a 10\% probability of being greater than 1.
    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$. The likelihood is a beta-binomial model (see discussion in \cref{transmission:sec:data}). an overdispersed binomial. 
\end{itemize}
All parameters have the default R-INLA priors, except where stated above.
For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.

@@

Having fitted the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the likelihood $\vec{P} \mid \vec{Z}$ is well-approximated by only considering its mean.
\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.

The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
This gives a lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
$\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
\begin{align}
Z_t
&= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
Unlike an inference procedure which infers the parameters simultaneously, there is no feedback loop: the value of $Z_{t}$ does not affect $Z_{t'}$ for $t' < t$.
The second term here is the contribution of the incidence prior to day $t$ to day $t$'s prevalence.
The remainder of the prevalence on day $t$ is therefore due to incidence on day $t$.
% Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
% Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.

Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
There are two sets of assumptions which are most likely to be violated.
The first set of assumptions is in constructing survival matrix, $\matr{S}$, which is assumed to be known with no uncertainty in this procedure.
However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.

A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
\todo{Maybe move this paragraph to the discussion?}
Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
% MCMC is likely too computationally expensive.
% Other approaches, such as a Laplace approximation or variational inference, could be considered.
For this analysis, I replaced any value of incidence estimated to be negative with 0.

@@

@@

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs.
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information for the posterior to overcome the prior penalisation towards this situation.
The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

@@

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows a infected, pre-positive compartment, which I denote $p_0$, to be included simply.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
